Modeling workflow: Lake $N_2O$ survey data

### 2023-03-14 scripts/NLA_N2O_models.Rmd

---
title: "Modeling workflow: Lake $N_2O$ survey data"
author: "Roy Martin, Jake Beaulieu, Michael McManus"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: TRUE
    toc_depth: 4
    number_sections: TRUE
    df_print: "tibble"
    math_method: 
      engine: webtex
      url: https://latex.codecogs.com/svg.image?
    html_preview: TRUE
    keep_html: TRUE
bibliography: RWM_Endnote_Library.bib
link-citations: yes
editor_options:
  chunk_output_type: inline
---

@@

df_model %>%
  print()
```

@@

df_test %>%
  print()
```

@@

sframe %>%
  print()
```

@@

sframe %>%
  group_by(WSA9, state) %>%
  summarise(n = n(), .groups = "drop") %>%
  spread(state, n) %>%
  print()
```

@@

sframe %>%
  group_by(WSA9, state, size_cat) %>%
  summarise(n = n(), .groups = "drop") %>%
  spread(size_cat, n) %>%
  print()
```

@@

pframe %>%
  print()
```

@@

samp_props %>%
  print()
```

@@

In the $\textbf{brms}$ package, there is additionaly functionality and syntax for multivariate responses and for allowing the varying intercepts in a multivariate model to be correlated, e.g.,:

$$ N_2O_{diss}\sim 1 + (1|a|WSA9) + (1|b|WSA9:state) + (1|c|WSA9:state:size) \\ 
N_2O_{equi}\sim 1 + (1|a|WSA9) + (1|b|WSA9:state) + (1|c|WSA9:state:size)$$

@@

To make inferences from this model for N2O in the population of interest, however, the included covariates needed to be (1) fully observed across that population or (2) their missingness needed to be modeled. For the lake area and elevation covariates, data _was_ available for all lakes from previously compiled geospatial databases. However, neither surface temperature or NO3 were observed for lakes outside of the sample. They were partially observed with respect to the target population. Their missingness needed to be accounted for in a model. Therefore, a more complex model was constructed below that included surface temperature and NO3 as additional responses conditioned on the survey design variables and fully observed covariates. This approach to inference for N2O was similar to a Bayesian structural equation model [@Merkle_etal_2021; @Merkle_Rosseel_2018]. The main details of the logical dependence structure could be characterized as:

$$\begin{align} 
\color{#1F449C}{\boldsymbol{N_2O_{diss}}} \sim Survey + Area + \color{#F05039}{\boldsymbol{NO_3}} + \color{#EEBAB4}{\boldsymbol{Temp}} \\ 
\color{#A8B6CC}{\boldsymbol{N_2O_{equil}}} \sim Survey + Elev  + \color{#EEBAB4}{\boldsymbol{Temp}}\\
\color{#F05039}{\boldsymbol{NO_3}} \sim Survey + Area + \color{#EEBAB4}{\boldsymbol{Temp}} \\ 
\color{#EEBAB4}{\boldsymbol{Temp}} \sim Survey + Lat + Elev + Day
\end{align}$$

Variables in color text above were treated as partially observed with respect to the population of interest (i.e., observed only in the sample), whereas variables in black text were considered fully observed. The partially observed variables, being dissolved and equilibrium N2O, NO3, and surface temperature, were each modeled conditional on the survey design variables and other partially and/or fully observed covariates. This structural equation approach requires a more complex set of post-processing steps compared to a typical MRP analysis. In order to propagate estimates and uncertainty through the dependency structure and make inferences, the fitted model was used to first predict surface temperature in the target population, since it depended only on the fully observed covariates. That predictive distribution was then used alongside the relevant fully observed covariates to predict NO3 in the target population. Finally, the predictive distributions for termperature and NO3 were used to predict the N2O responses. These steps were carried out in the "Predict to population" section to follow.

In the final model below, the submodel for surface temperature assumed a Gamma distributed error distribution and the linear predictor included the survey design variables, latitude, elevation, and julian date. The shape parameter was also modeled as a function of latitude to address increasing response variance along the latitudinal gradient. The NO3 submodel was a cumulative logit formulation and the linear predictor included all of the survey factors as well as surface temperature and lake area.

>>>>

---
title: "Modeling workflow: Lake $N_2O$ survey data"
author: "Roy Martin, Jake Beaulieu, Michael McManus"
date: "`r Sys.Date()`"
output:
  github_document:
    number_sections: TRUE
    df_print: "tibble"
    math_method: 
      engine: webtex
      url: https://latex.codecogs.com/svg.image?
    html_preview: TRUE
    keep_html: TRUE
bibliography: RWM_Endnote_Library.bib
link-citations: yes
editor_options:
  chunk_output_type: inline
---

@@

df_model %>%
  head(10)
```

@@

df_test %>%
  head(10)
```

@@

sframe %>%
  head(10)
```

@@

sframe %>%
  group_by(WSA9, state) %>%
  summarise(n = n(), .groups = "drop") %>%
  spread(state, n) %>%
  head(10)
```

@@

sframe %>%
  group_by(WSA9, state, size_cat) %>%
  summarise(n = n(), .groups = "drop") %>%
  spread(size_cat, n) %>%
  head(10)
```

@@

pframe %>%
  head(10)
```

@@

samp_props %>%
  head(10)
```

@@

In the $\textbf{brms}$ package, there is functionality and syntax for multivariate responses and for allowing the varying intercepts in a multivariate model to be correlated, e.g.,:

$$N_2O_{diss} \sim 1 + (1|a|WSA9) + (1|b|WSA9:state) + (1|c|WSA9:state:size) \\\ 
N_2O_{equi} \sim 1 + (1|a|WSA9) + (1|b|WSA9:state) + (1|c|WSA9:state:size)$$

@@

To make inferences from this model for N2O in the population of interest, however, the included covariates needed to be (1) fully observed across that population or (2) their missingness needed to be modeled. For the lake area and elevation covariates, data _was_ available for all lakes from previously compiled geospatial databases. However, neither surface temperature or NO3 were observed for lakes outside of the sample. They were partially observed with respect to the target population. Their missingness needed to be accounted for in a model. Therefore, a more complex model was constructed below that included surface temperature and NO3 as additional responses conditioned on the survey design variables and fully observed covariates. This approach to inference for N2O was similar to a Bayesian structural equation model [@Merkle_etal_2021; @Merkle_Rosseel_2018]. The logical dependence structure could be characterized as:

$$\begin{align} 
{\boldsymbol{N_2O_{diss}}} &\sim Survey + Area + {\boldsymbol{NO_3}} + {\boldsymbol{Temp}} \\ 
{\boldsymbol{N_2O_{equil}}} &\sim Survey + Elev  + {\boldsymbol{Temp}}\\
{\boldsymbol{NO_3}} &\sim Survey + Area + {\boldsymbol{Temp}} \\ 
{\boldsymbol{Temp}} &\sim Survey + Lat + Elev + Day
\end{align}$$

Variables in bold text above were treated as partially observed with respect to the population of interest (i.e., observed only in the sample), whereas variables not in bold were considered fully observed. The partially observed variables, being dissolved and equilibrium N2O, NO3, and surface temperature, were each modeled conditional on the survey design variables and other partially and/or fully observed covariates. This structural equation approach requires a more complex set of post-processing steps compared to a typical MRP analysis. In order to propagate estimates and uncertainty through the dependency structure and make inferences, the fitted model was used to first predict surface temperature in the target population, since it depended only on the fully observed covariates. That predictive distribution was then used alongside the relevant fully observed covariates to predict NO3 in the target population. Finally, the predictive distributions for temperature and NO3 were used to predict the N2O responses. These steps were carried out in the "Predict to population" section to follow.

In the final model below, the sub-model for surface temperature assumed a Gamma distributed error distribution and the linear predictor included the survey design variables, latitude, elevation, and Julian date. The shape parameter was also modeled as a function of latitude to address increasing response variance along the latitudinal gradient. The NO3 sub-model was a cumulative logit formulation and the linear predictor included all of the survey factors as well as surface temperature and lake area.
